Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-06 aistats/paper.tex
Remove mixed term from biasvariance figures
attempt at clarification for intro

We kind of know how to do optimization 
on smooth, strongly convex functions 
when there is some bound on the stochasticity,
for example in the form of a bounded variance, 
say for the problem 
\begin{equation}
\min_\theta f(\theta)
\end{equation}
with access to stochastic gradients 
$g(\theta)$ such that 
\begin{equation}
	\expect{\norm{g(\theta) - \nabla f(\theta)}^2} \leq \sigma^2.
\end{equation}
The best convergence rate, of stochastic gradient methods 
is obtained by SGD with decreasing step-sizes and averaging, 
yields a convergence rate of $1/t$ and is asymptotically tight;
SGD achieves this rate on the following problem, 
\begin{equation}
	f(\theta) = \frac{1}{2}\norm{\theta -\mu}^2, \quad
	g(\theta) \sim \sim \cN(\theta - \mu, \sigma^2).
\end{equation}
This problem is estimating the mean of a Gaussian, 
and it is not improvable by information theoretic arguments.

But our understanding of stochastic optimization lacks in other domains. 
A generalization of the above problem, 
at the intersection of optimization and estimation, 
is the problem of minimizing 
\begin{equation}
	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{A(\theta) - \lin{t(x), \theta}},
\end{equation}
where $A$ is a convex, Legendre function and 
$p(x\cond\theta)$ is the exponential family distribution 
with $A$ as its log-partition function. 
The Gaussian example above is a special case 
with $A(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$
and $p(x\cond\theta) = \cN{\mu, \sigma^2}$.

We obviously have algorithms for this setting, 
as this is an maximum likelihood problem.
Giving the (online) maximum likelihood after seeing $t$ samples
which corresponds as online mirror descent with a decreasing step-size schedule (more on this later).
But our convergence rates are weak.
There are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.

We do not have a solution. 
The goal of this paper, and the reason it is submitted at AISTATS, 
is to bring this issue to the attention of 
the community working at the intersection of machine learning, optimization and statistics community,
in the hope that we can make progress on this question;
Can we build algorithms for stochastic estimation 
when the objective function is not (close to) quadratic
and the noise is not (sub-)Gaussian?

We give motivating examples where solution of this problem would enable progress in other areas,
principally motivated by stochastic optimization, 
show the connections between the optimization and statistical estimation sides of the problem,
and discuss where our current theory fails to fully capture the problem.

We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate---this is part of the point of this paper---and that the statistical theory results for the convergence of the MLE to the solution 
are well known. 
Our point is more subtle, in that for the purposes of optimization, 
to mimic the results that are known for SGD in Euclidean settings, 
we would like a result of the form 
\begin{equation*}
	\text{optimality gap after $t$ steps}
	\leq 
	\frac{\text{``bias''}}{t^2}
	+ \frac{\text{``variance''}}{t},
\end{equation*}
which is the bit that is lacking.

@@

\begin{figure*}[ht]
	\centering
	\includegraphics[width=\textwidth]{{figs/biasvariance/linear_ratio=0.8_n0=10}.pdf}
	\caption{
	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound. 
	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
	}
	\label{fig:variance_decomposition}
\end{figure*}

>>>>

We kind of know how to do optimization 
on smooth, strongly convex functions 
when there is some bound on the stochasticity,
for example in the form of a bounded variance, 
say for the problem 
\begin{equation}
\min_\theta f(\theta)
\end{equation}
with access to stochastic gradients 
$g(\theta)$ such that 
\begin{equation}
	\expect{\norm{g(\theta) - \nabla f(\theta)}^2} \leq \sigma^2.
\end{equation}
Estimating the mean of Gaussian noise is a special case of this problem class, using 
$f(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$ and $g(\theta) \sim \cN(\theta - \mu, \sigma^2)$.
SGD with a decreasing step-size coincides with the maximum likelihood estimate (MLE), 
and attains the $1/t$ convergence rate.

But our understanding of stochastic optimization lacks in other domains. 
A generalization of the above problem, at the intersection of optimization and statistics, 
is the problem of minimizing 
\begin{equation}
	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta_*)},
\end{equation}
When $p(x\cond\theta)$ is an exponential family distribution,
$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (up to constants).
Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
The Gaussian example is the special case $A(\theta) = \frac{1}{2}\norm{\theta}^2$.

There are algorithms for this setting, for example the MLE;
giving the (online) maximum likelihood after seeing $t$ samples
corresponds to online mirror descent with decreasing step-sizes (more on this later).
But our convergence rates are weak; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.

We do not have a solution. The goal of this paper, and the reason it is submitted at AISTATS, 
is to bring this issue to the attention of the community working at the intersection of machine learning, optimization and statistics,
in the hope that we can make progress on this question;
Can we build algorithms for stochastic estimation 
when the objective function is not (close to) quadratic
and the noise is not Gaussian?

We show the connections between the optimization and statistical estimation view of the problem
and discuss where our current theory fails to fully capture the problem.
We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate---this is part of the point of this paper---and 
that the statistical theory results for the convergence of the MLE to the solution are well known. 
Our point is more subtle, in that for the purposes of optimization, to mimic the results that are known for SGD in Euclidean settings, 
we would like a result of the form 
\begin{equation*}
	\text{optimality gap after $t$ steps}
	\leq 
	\frac{\text{``bias''}}{t^2}
	+ \frac{\text{``variance''}}{t},
\end{equation*}
which is the bit that is lacking.

@@

\begin{figure*}[ht]
	\centering
	\includegraphics[width=\textwidth]{{figs/biasvariance/new_linear_ratio=0.8_n0=10}.pdf}
	\caption{
	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound. 
	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
	}
	\label{fig:variance_decomposition}
\end{figure*}
